Prisoner's Dilemma

### 2020-05-06 realnotes2/notes.tex
almost finish proof of theorem1

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{problem}{Problem}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{example}{Example}
\newtheorem*{solution}{Solution}

@@

Lots to add here.

@@

\begin{definition}
  \label{strategypayoffs}
  Let $X_t$ be the random variable designating which state the $s_1$-$s_2$ Markov chain is in at time $t$. Then, the \text{payoff} of strategy $s_1$ when played against strategy $s_2$ is 
  \begin{equation*}
    v_{s_1}(s_2) = E \left[ \lim_{T \to \infty} \frac{1}{T} \sum_{t=0}^T r(G_{s_1,s_2} (X_t)) \Bigm| X_0 = (c_{\text{start}}(s_1), c_{\text{start}}(s_2)) \right]
  \end{equation*}
\end{definition}

@@

\begin{proof}
  The key idea is that a time average sum where each element is one of finitely many values can be written as a frequency-weighted finite sum instead. Let $I_{c_1,c_2,t}$ be the indicator variable that is 1 if $G_{s_1,s_2}(X_t) = (c_1,c_2)$ and 0 otherwise.  Then, we can write \begin{equation*}
    \lim_{T \to \infty} \frac{1}{T} \sum_{t = 0}^T r(G_{s_1,s_2}(X_t)) = 
    \lim_{T \to \infty}\frac{1}{T} \sum_{t = 0}^T \sum_{(c_1,c_2) \in S_{s_1,s_2}} r(c_1,c_2) \cdot I_{c_1,c_2,t}
  \end{equation*}
  We may now exchange the order of summation and move the finite sum out of the limit, to get \begin{equation*}
    \lim_{T \to \infty}\frac{1}{T} \sum_{t = 0}^T \sum_{(c_1,c_2) \in S_{s_1,s_2}} r(c_1,c_2) \cdot I_{c_1,c_2,t}
    = \sum_{(c_1,c_2) \in S_{s_1,s_2}} r(c_1,c_2) \cdot \lim_{T \to \infty} \sum_{t=0}^T \frac{I_{c_1,c_2,t} }{T}
  \end{equation*}
  We can now use \cref{strategypayoffs} and linearity of expectation to find that
  \begin{equation*}
    v_{s_1}(s_2) = \sum_{(c_1,c_2) \in S_{s_1,s_2}} r(c_1, c_2) E \left[ \lim_{T \to \infty} \sum_{t=0}^T \frac{I_{c_1,c_2,t}}{T} \bigm| X_0 = (c_{\text{start}}(s_1), c_{\text{start}}(s_2))
    \right]
  \end{equation*}
  Finally, we note that this is exactly the statement of \cref{payofftimeaverage}, which proves our lemma.
\end{proof}

@@

\begin{figure}
  \label{tftfigure}
  \includegraphics[width=4cm]{tft.jpg}
  \centering
  \caption{TFT.}
\end{figure}
\begin{figure}
  \label{pavlovfigure}
  \includegraphics[width=4cm]{pavlov.jpg}
  \centering
  \caption{Pavlov.}
\end{figure}

\begin{remark}
  Tit-for-tat, displayed in \cref{tftfigure}, is not evolutionarily stable. It has the stationary distribution $(1/4,1/4,1/4,1/4)$ in its own Markov chain, which has a payoff that is significantly smaller than $R$. This should be intuitive: if tit-for-tat makes one mistake, it goes into a defection cycle that it doesn't break out of until it makes a second mistake.
\end{remark}

@@

\begin{lemma}
  \label{limitvexists}
  The limit \begin{equation*}
    \lim_{p \to 0} v_{s_1}(s_2)
  \end{equation*}
  exists, for any strategies $s_1$ and $s_2$.
\end{lemma}
\begin{proof}
  Hmmmm this was harder than I expected.
  \todo{prove this; seems obvious}
\end{proof}

@@

      We create the strategy $s_2$ as follows. First, copy the entire $s_1$ machine into $s_2$. Suppose that the state corresponding to the start state of $s_1$ is $c_s$. Recall that the output at $c_s$ is $G(c_s)$, and that the state $s$ goes to upon perceiving the opponent move $G(c_s)$ is $T(c_s, G(c_s))$. Now, create two new states: $c_0$ and $c_1$. Define the transitions as \begin{align*}
        T(c_0, G(c_s)) &= T(c_s, G(c_s))\\
        T(c_0, \lnot G(c_s)) &= c_1 \\
        T(c_1, \cdot) &= c_1
      \end{align*}
      and the outputs as \begin{align*}
        G(c_0) &= \lnot G(c_2)\\
        G(c_1) &= C.
      \end{align*}
      Let the start state of $s_2$ be $c_0$.

      Given this construction, we claim that the rewards are as follows:
      \begin{align*}
        v_{s_1}(s_1) &= \gamma \\
        v_{s_1}(s_2) &\leq (1-p) \gamma + p T \\
        v_{s_2}(s_1) &\geq (1 - p) \gamma  + p S \\
        v_{s_2}(s_2) &\geq (1-p)^{2} R + 2 (1-p) p (\tfrac{S + T}{2}) + p^2 \gamma 
      \end{align*}
      \todo{prove this! using time-average distributions this becomes relatively trivial}

      Now, we simply compute $F(s_2) - F(s_1)$, which we want to show is greater than 0.

      \begin{align*}
        F(s_2) - F(s_1) &= \\
        &= (1 - \epsilon) \cdot v_{s_2}(s_1) + \epsilon \cdot v_{s_2}(s_2) - (1 - \epsilon) \cdot v_{s_1}(s_1) - \epsilon \cdot v_{s_1}(s_2) \\
        &= (1 - \epsilon) (\gamma + p(\ldots)) + \epsilon (R + p(\ldots)) - (1-\epsilon) \gamma - \epsilon (\gamma + p(\ldots)) \\
        &= \epsilon (R - \gamma) + p(\ldots)
      \end{align*}

      $R - \gamma > 0$ by our assumption. Clearly, since $(\ldots)$ is some polynomial in $p$, given an $\epsilon$ we can find a sufficiently small $p$ such that the full expression is positive. This proves that $s_2$ can invade $s_1$, and thus, that $s_1$ is not $\epsilon$-invadable for this value of $p$. In conclusion, then $s_1$ is not evolutionarily stable, which concludes the proof of \cref{evolutionarystable1}.

@@

>>>>

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{claim}{Claim}
\newtheorem{problem}{Problem}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{example}{Example}
\newtheorem*{solution}{Solution}

@@

Introduce the prisoner's dilemma, the iterated prisoner's dilemma and the goal of finding out how cooperation can appear. Introduce Axelrod along with some other previous results.

@@

\begin{definition}
  \label{strategypayoffs}
  Let $X_t$ be the random variable designating which state the $s_1$-$s_2$ Markov chain is in at time $t$. The \text{payoff} of strategy $s_1$ when played against strategy $s_2$ is 
  \begin{equation*}
    v_{s_1}(s_2) = E \left[ \lim_{T \to \infty} \frac{1}{T} \sum_{t=0}^T r(G_{s_1,s_2} (X_t)) \Bigm| X_0 = (c_{\text{start}}(s_1), c_{\text{start}}(s_2)) \right]
  \end{equation*}
\end{definition}

@@

We may also make $r(c_1,c_2)$ into a vector, denoted by $r$, and write this as \begin{equation*}
  v_{s_1}(s_2) = \pi^{(c_{\text{start}}(s_1), c_{\text{start}}(s_2))} \cdot r.
\end{equation*}

\begin{proof}[Proof of \cref{payofftimeaverage}]
  The key idea is that a time average sum where each element is one of finitely many values can be written as a frequency-weighted finite sum instead. Let $I_{c_1,c_2,t}$ be the indicator variable that is 1 if $G_{s_1,s_2}(X_t) = (c_1,c_2)$ and 0 otherwise.  Then, we can write \begin{equation*}
    \lim_{T \to \infty} \frac{1}{T} \sum_{t = 0}^T r(G_{s_1,s_2}(X_t)) = 
    \lim_{T \to \infty}\frac{1}{T} \sum_{t = 0}^T \sum_{(c_1,c_2) \in S_{s_1,s_2}} r(c_1,c_2) \cdot I_{c_1,c_2,t}
  \end{equation*}
  We may now exchange the order of summation and move the finite sum out of the limit, to get \begin{equation*}
    \lim_{T \to \infty}\frac{1}{T} \sum_{t = 0}^T \sum_{(c_1,c_2) \in S_{s_1,s_2}} r(c_1,c_2) \cdot I_{c_1,c_2,t}
    = \sum_{(c_1,c_2) \in S_{s_1,s_2}} r(c_1,c_2) \cdot \lim_{T \to \infty} \sum_{t=0}^T \frac{I_{c_1,c_2,t} }{T}
  \end{equation*}
  We can now use \cref{strategypayoffs} and linearity of expectation to find that
  \begin{equation*}
    v_{s_1}(s_2) = \sum_{(c_1,c_2) \in S_{s_1,s_2}} r(c_1, c_2) E \left[ \lim_{T \to \infty} \sum_{t=0}^T \frac{I_{c_1,c_2,t}}{T} \bigm| X_0 = (c_{\text{start}}(s_1), c_{\text{start}}(s_2))
    \right]
  \end{equation*}
  Finally, we note that this is exactly the statement of \cref{payofftimeaverage}, which proves our lemma.
\end{proof}

@@

\begin{figure}
  \label{tftfigure}
  \includegraphics[width=4cm]{tft.jpg}
  \centering
  \caption{TFT.}
\end{figure}
\begin{figure}
  \label{pavlovfigure}
  \includegraphics[width=4cm]{pavlov.jpg}
  \centering
  \caption{Pavlov.}
\end{figure}
  \todo{use tikz for these figures!}

\begin{remark}
  Tit-for-tat, displayed in \cref{tftfigure}, is not evolutionarily stable. It has the stationary distribution $(\frac{1}{4},\frac{1}{4},\frac{1}{4},\frac{1}{4})$ in its own Markov chain, which has a payoff that is significantly smaller than $R$. This should be intuitive: if tit-for-tat makes one mistake, it goes into a defection cycle that it doesn't break out of until it makes a second mistake.

\end{remark}

@@

\begin{lemma}
  \label{limitvexists}
  The limit \begin{equation*}
    \lim_{p \to 0} v_{s_1}(s_2)
  \end{equation*}
  exists, for any strategies $s_1$ and $s_2$.
\end{lemma}
\begin{proof}
  Use the fact that this is the dot product of a stationary distribution with a fixed coefficient vector, and then say something about how any stationary distribution with the correct sum needs to be a polynomial in $p$ maybe?
  \todo{prove this; seems obvious}
\end{proof}

@@

      We create the strategy $s_2$ as follows. First, copy the entire $s_1$ machine into $s_2$. Suppose that the state corresponding to the start state of $s_1$ is $c_s$. Recall that the output at $c_s$ is $G(c_s)$, and that the state $s$ goes to upon perceiving the opponent move $G(c_s)$ is $T(c_s, G(c_s))$. Now, create two new states: $c_0$ and $c_1$. Define the transitions as \begin{align*}
        T(c_0, G(c_s)) &= T(c_s, G(c_s))\\
        T(c_0, \lnot G(c_s)) &= c_1 \\
        T(c_1, \cdot) &= c_1
      \end{align*}
      and the outputs as \begin{align*}
        G(c_0) &= \lnot G(c_2)\\
        G(c_1) &= C.
      \end{align*}
      Let the start state of $s_2$ be $c_0$. \todo{add a figure!! the construction is simple but this description is deceivingly hard}

      \begin{claim}
        \label{claimpayoffs}
        Given the above construction of $s_2$, the following inequalities hold:
      \begin{align*}
        v_{s_1}(s_1) &\leq (1-p)^2 \gamma + 2(1-p)p R + p^2  R \\ 
        v_{s_1}(s_2) &\leq (1-p) \gamma + p T \\
        v_{s_2}(s_1) &\geq (1 - p) \gamma  + p S \\
        v_{s_2}(s_2) &\geq (1-p)^{2} R + 2 (1-p) p (\tfrac{S + T}{2}) + p^2 \gamma .
      \end{align*}
      \end{claim}

      We will first finish the proof of the theorem assuming this claim, and then prove it.

      Now, we simply compute $F(s_2) - F(s_1)$, which we want to show is greater than 0.
      \begin{align*}
        F(s_2) - F(s_1) &= \\
        &= (1 - \epsilon) \cdot v_{s_2}(s_1) + \epsilon \cdot v_{s_2}(s_2) - (1 - \epsilon) \cdot v_{s_1}(s_1) - \epsilon \cdot v_{s_1}(s_2) \\
        &= (1 - \epsilon) (\gamma + p(\ldots)) + \epsilon (R + p(\ldots)) - (1-\epsilon) (\gamma + p(\ldots)) - \epsilon (\gamma + p(\ldots)) \\
        &= \epsilon (R - \gamma) + p(\ldots)
      \end{align*}

      We know that $R - \gamma > 0$ by our initial assumption. Clearly, since $(\ldots)$ is some polynomial in $p$, given an $\epsilon$ we can find a sufficiently small $p$ such that the full expression is positive. This proves that $s_2$ can invade $s_1$, and thus, that $s_1$ is not $\epsilon$-invadable for this value of $p$. In conclusion, then $s_1$ is not evolutionarily stable, which concludes the proof of \cref{evolutionarystable1}.

@@

    \begin{proof}[Proof of \cref{claimpayoffs}]
      We can prove this using either of the two definitions.
    \end{proof}
